---
title: "Difference between R and SAS in estimating median survival time"
author: "Zhanwen Peng"
date: "2025-01-03"
categories: [R, SAS, survival]
code-block-border-left: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
Sys.setLanguage('en')
Sys.setlocale("LC_TIME", "US.UTF-8")
```

```{r, message=FALSE}
library(SASmarkdown)
library(tidyverse)
library(survival)
library(survminer)
```

In this post we are going to discuss about the difference between R and SAS in estimating median survival time.

Firstly, Create a simple survival dataset in tibble form.

```{r}
surv_data <- tribble(
  ~AVAL, ~CNSR,
  1,   0,
  4,   0,
  7,   1,
  8,   1,
)
surv_data
```

Using `survival` package in `r` to estimate the median survival time and plot the KM graph.

```{r}
km_model <- survfit(Surv(AVAL, CNSR==0) ~ 1, data = surv_data,conf.type=c("log-log"))
ggsurvplot(km_model, data = surv_data, risk.table = FALSE, conf.int = F,
           title = "Dummy survival data", 
           submain = "Kaplan-Meier estimates", 
           break.x.by = 1,
           ylab = "Survival",
           xlab = "AVAL",
           legend.title = "",surv.median.line = "hv")
```

we can see that in $t=4$, the KM graph coincides with line $y=0.5$, then we estimate the median survival time in `r`,

```{r}
quantile(km_model,0.5)$quantile
```

According the result above, the estimated median is $6$.

What if we use the same dataset but estimate survival median using `proc lifetest` in `SAS` instead.
```{sas engine="sas"}
option locale='en_us';
data surv_data;
	input AVAL CNSR;
	cards;
	1 0
	4 0
	6 1
	8 1
	;
run;

ods exclude all;
ods output productlimitestimates=est Quartiles=ql;
proc lifetest data=surv_data method=km reduceout outsurv=plc ;
    time AVAL*CNSR(1);
run;
ods exclude close;

proc print data=ql;
run;
```
We can see that the estimation of median is actually `NA`, this means that `R` and `SAS` are using different formula to estimate the median.

In short, `survival` package in `R` and early version of `SAS` (such as 9.1) estimate the quantiles using:
$$
\begin{aligned}
q_p &= \frac{1}{2}(\inf_t\{1-\hat{S}(t)\ge p\}+\sup_t\{1-\hat{S}(t)\le q\})\\
&=\frac{1}{2}(\inf_t\{\hat{S}(t)\le 1-p\}+\sup_t\{\hat{S}(t)\ge 1-q\})\
\end{aligned}
$$
Therefore the result from `survival` is $\frac{1}{2}(4+8)=6$.

In `SAS 9.4`, the formula is
$$
q_p = \min\{t_j|\hat{S}(t_j)<1-p\}
$$
Since there is no data point such that $\hat{S}(t_j)$ is strictly less then $0.5$, the estimation of median from `SAS` is `NA`.

### References
[proc lifetest](https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.4/statug/statug_lifetest_details03.htm)\

[survival: Survival Analysis](https://cran.r-project.org/web/packages/survival/survival.pdf)

